The \dealii{} Library, Version 9.2

### 2020-06-01 9.2/paper.tex
Add section on SD::BatchOptimizer

\todo[inline]{J-P's section}

BatchOptimizer for symbolic expressions
\begin{itemize}
  \item CSE for dictionary-based expressions
  \item ``lambdify'' (via SymEngine)
  \item Offloading to LLVM JIT compiler (via SymEngine)
  \item Serialization, so complex expressions can be compiled offline
\end{itemize}

>>>>

In the previous release we added support for symbolic expressions, leveraging the
SymEngine library \cite{symengine-web-page}.
Although effective, evaluating lengthy expressions could be a bottle-neck as this
was performed using dictionary-based substitution.
We have improved on this by implementing a \texttt{BatchOptimizer} class in the 
namespace \texttt{Differentiation::SD} that collects several \texttt{Expression}s
and transforms them in such a way that the equivalent result is returned through
a quicker code path.
This may be done by simply using common subexpression elimination (CSE) for the
dictionary-based expressions, by transformation to a set of nested 
\texttt{std::function}s (the equivalent to \texttt{SymPy}'s "lambdify", with or
without using CSE), or by offloading to these expressions to the \texttt{LLVM}
just-in-time (JIT) compiler.
Although, each of these features is implemented and tested in the SymEngine
library itself, the \texttt{BatchOptimizer} class provides both uniform
interface to their classes and a convenient interface for scalar expressions,
as well as tensorial expressions formed using the \texttt{deal.II} tensor and 
symmetric tensor classes.
It, like the \texttt{Expression} class, is also serializable.

The way that the batch optimizer may be employed within a user's code is shown
in the pseudo-code below.
As per usual, one would first define some independent variables, and 
subsequently compute some symbolic expressions that are dependent on these 
independent variables. 
These expression could be, for example, scalar  expressions or tensors of 
expressions.
Instead of evaluating these expressions directly, the user would now create an
optimizer to evaluate the dependent functions.
In this example, the selected arithmetic type numerical result will be of type
\texttt{double}, and the \texttt{LLVM} JIT optimizer will be invoked.
It will employ common subexpression elimination and aggressive optimizations
during compilation.
The user then informs the optimizer of all of the independent variables and the
dependent expressions, and invokes the optimization process.
This is an expensive call, as it determines an equivalent code path to evaluate
all of the dependent functions at once.
However, in many cases each evaluation has significantly less computational cost
than evaluating the symbolic expressions directly.
Evaluation is performed when the user constructs a substitution map, giving each
independent variable a numerical representation, and passes those to the 
optimizer.
After this step, the numerical equivalent of the individual dependent expressions
may finally be retreived from the optimizer.

\begin{c++}
using namespace Differentiation::SD;

const Expression x("x");
const Expression y("y");
...
const auto f = calculate_f(x, y, ...); // User function
const auto g = calculate_g(x, y, ...); // User function
...

BatchOptimizer<double> optimizer (OptimizerType::llvm,
                                  OptimizationFlags::optimize_all);
optimizer.register_symbols(x, y, ...);
optimizer.register_functions(f, g, ...);
optimizer.optimize();

const auto substitution_map
  = make_substitution_map({x, ...}, {y, ...}, ...);
optimizer.substitute(substitution_map);

const auto result_f = optimizer.evaluate(f);
const auto result_g = optimizer.evaluate(g);
\end{c++}

This expense of invoking the optimizer may be offset not only by
the number of evaluations performed, but also by the amount of reuse each
instance of a \texttt{BatchOptimizer} has.
In certain circumstances, this can be maximized by generalizing the way in which
the dependent expressions are formulated.
For example, in the context of constitutive modelling the material coefficients
may be made symbolic rather than encoding these into the dependent expressions
as numerical values.
The optimizer may then be used to evaluate an entire family of constitutive laws,
and not a specific one that describes the response of a single material.
Thereafter, serializing the optimizer instance and reloading the contents during
subsequent simulations permits the user to skip the optimization process entirely.
Serialization also enables these complex expressions to be compiled offline.
